Comparing Power Optimizations in the Reorder Buffer

### 2011-06-08 paper/paper.tex
multi file paper

\maketitle
\begin{abstract}
	In an effort to increase instruction level parallelism techniques such 
as pipelining, out of order execution and operand forwarding.  However 
these efforts come at a large power cost.  Out of order execution requires
a ReOrder Buffer (ROB), to commit the registers in program order.  This 
structure has grown to be monolithic with numerous entries and read/write ports.
There is no current methods to efficiently and equally compare various
ROB advancements to improve power efficiency. We propose a loosely clycle
accurate model that keeps track of the power usages of the ROB.  Using the 
simulation tool, we are able to compare and contrast the affects of improvements
made on the ROB.  This also allows for a first line test to validate proof of 
concepts for future power efficient ROB designs. 
\end{abstract}

\terms{Power, Energy, Architecture}

\keywords{Power efficiency, ROB, Simulation} % NOT required for Proceedings

\section{Introduction}
BLAH BLAH BLAH INTRO WEEEE
\section{Background}
    An out of order execution system has two additional stages 
to the lifetime of the instruction on top of the traditional pipeline 
stages.  Each of which a small memory array to keep track of instuction flow.
The first addtion, the instruction queue(IQ) which holds instructions that are 
waiting to be executed.  Each entry will hold the instruction -- including 
type, operand registers, and destination registers.  The second structure 
holds the instructions that are in flight but not yet commited to the register
file. Each entry will hold both PC of the instruction, as well as the destination
register and the result of the excution unit.  There is also a flag that will
mark the result of the entry as valid.

Replacing the traditional instruction dispatch stage, An placeholder will be made for the 
instruction in ReOrder Buffer (ROB), and the instructions will be placed
in the instruction queue(IQ). In this stage instructions will first wait for their operands if not already
available.  When all operands are available, the instruction is ready to 
be executed so wait until its appropriate execution unit is free.  At this 
point the instruction exits the IQ and begins execution.

Upon execution completion, the output value is stored in the ROB with its 
corresponding entry and that entry is marked as valid or completed with a 
1 bit flag.  An entry is commited from the ROB to the architectrual register
file (ARF) when that entry is the oldest (lowest PC) and its result has been 
marked as valid. 
\subsection{Circular ROB}
The circular ROB is 
\subsection{Distributed ROB}
An accepted optimization for improving the power consumption of an ROB unit 
is to break up the ROB buffer into clusters REF[XXX].  In this organization, the in-flight 
instruction information is distributed across all of the units and then re-assembled 
back into program order after the fact.  The primary advantage of this approach 
is that it dramatically lessens the size of the individual ROB units.  This reduces overall
power by reducing the bit line width as well as unit capacitance.

...IMAGE FROM OUR SLIDES...

The overhead associated with this scheme is that the addition of a separate FIFO unit
is required to maintain a global state of instruction order between the separate ROB units.
This central FIFO unit is then referenced upon instruction re-order to commit the instructions
in program order.  The contents of the FIFO unit are fairly simple and incur reasonable hardware
overhead.  Each entry needs only contain the PC of the instruction (or some other unique identifier)
and the corresponding ROB unit that houses the placeholder of the instruction as it is in-flight. 
The total ROB bit storage count remains the same from the non-distributed model to the distributed model
as the overall size of the ROB remains the same.  However, the inclusion of the central FIFO queue 
induces some overhead bits that must be considered when analyzing the overall effectiveness of the 
model.

The distribution or clustering of the ROB unit also inherently induces a minimal performance overhead 
in particular cases.  By breaking the ROB into smaller pieces and distributing the instruction 
data across the unit, the placement of instructions becomes a performance factor.  Typically, a 
distributed ROB model features only a single write port per ROB cluster.  As a result only one operand 
can be forwarded at a time from the ROB.  In the rare case that an instruction requires two operands be 
forwarded from the same ROB cluster a delay of one cycle is imposed while the operands are forwarded 
sequentially.  It is reported, however, that this happens very rarely in practice and induces less than 1\% 
performance degradation REF[XXX].
\subsection{Dynamic ROB}
\subsection{Latch ROB}

\section{Project Implementation}
Our simulator provides a cycle accurate behavioral model of a standard circular re-order buffer.  At the outset of the project, the standard re-order buffer we intended to implement was a non-circular RAM-based re-order buffer.  However, after feedback from the course instructor and students it was decided that a circular buffer is more widely held as being the 'standard' buffer type and would serve as an ideal underlying model for power optimizations.  Thus, we implement and compare the standard circular re-order buffer to circular re-order buffers with the following optimizations: 1) Distributed (Cluster based) ROB, 2) Dynamically Sized ROB, 3) Reduced Output Port ROB (Retention Latch). The following sections outline each of the above model implementations as well as input instruction generation and power results tabulation.  
\subsection{Instruction Generation}

\begin{table*}
\centering
\begin{tabular}{|c|c|c|c|} \hline
Instruction Type&Cycle Count&Instruction Type&Cycle Count\\ \hline
\texttt{LOAD}&20&\texttt{STORE}&20\\ \hline
\texttt{ADD}&1&\texttt{FADD}&3\\ \hline
\texttt{SUB}&1&\texttt{FSUB}&3\\ \hline
\texttt{MULT}&3&\texttt{FMULT}&5\\ \hline
\texttt{DIV}&18&\texttt{FDIV}&6\\ \hline\end{tabular}
\caption{Modeled instruction types and their corresponding completion cycle counts.}
\label{tab:instypetable}
\end{table*}

We model the input to our simulator as a stream of integer and floatpoint instructions.  The instructions are represented at the bit-level and model MIPS instructions as closely as possible.  Each instruction consists of 1) Instruction type, 2) Source and destination registers 3) a PC.  Table ~\ref{tab:instypetable} indicates the types of instructions modeled.  Each instruction type also has a predetermined cycle execution time designating the number of cycles that must elapse between instruction issue and the completion of the instruction.  As modeling functional units goes beyond the scope of this simulator, instructions are flagged as complete after the listed number of cycles has elapsed.  The actual cycle counts are the same as those modeled in MIT's Graphite many-core processor simulator\cite{graphite}, with the exception of LOAD and STORE.  The cycle counts corresponding to LOAD and STORE instructions are estimated as a predetermined 20 cycles each.  This is simply an estimation as the actual cycle count is heavily dependent on cache architecture and benchmark specific cache behavior.  Since both of these topics are beyond the scope of the simulator, we estimate the completion time of LOAD and STORE instructions as a set value substantially longer than the other more accurate instruction types.

Registers are generated in bit representation as 4 bits each totalling 12 bits for the s1, s2 and d registers in each instruction. Currently, our representation of instructions make no distinction between zero, 1, and 2 operand instructions.  All instructions regardless of type feature two operands.  This feature is supported by the fact that a memory hierarchy and functional units are not simulated by our system and as such immediate values have no direct use.  Further, every instruction features a d register regardless of type.  Because our simulator features operand forwarding from the ROB, it is necessary for the generation of registers to keep track of past destination registers to use as dependent operands in future instructions.  This history buffer is implemented as a rotating bit register from which prior destination registers can be indexed.

Each instruction also features a program counter associated with the instruction.  While this is clearly a simplification of a real system, it serves the purposes of our simulation as a uniquely identifiable tag by which to differentiate instructions.  In both a real system and our simulator, a statically set 32 bit value is used to represent this PC/tag.  As such, the specific values contained in the PC/tag will yield accurate transition results (nonexistent) and is thus are irrelevant.  Simply starting the PC of the first instruction in the stream at 0 and incrementing by 4 each instruction guarantees uniqueness among PCs and serves the purposes of an instruction tag.

The aforementioned instruction fields have multiple modes of generation.  Our simulator synthetically generates an instruction input trace based on a prescribed instruction profile.  The instruction profile features: 1) a percent rate of occurrence for each instruction type, 2) a rate corresponding to the interdependence of registers between instructions, and 3) an overall instruction count.

The instruction generator is currently capable of generating four distinct instruction streams corresponding to different instruction workloads.  The first instruction stream is marked as the default instruction stream.  It features equal percentages of each type of instruction as well as a 10% register dependence rate.  This register dependence rate was selected arbitrarily to act as a starting point for modeling data dependencies between instructions.

The remaining three instruction streams are modeled after SPEC-FP benchmarks \cite{rupnow}.  The benchmarks were chosen to represent three major modes of operation:  floating point heavy execution, integer heavy execution, and mixed type execution.  For a floating point heavy execution, 173.applu was selected and modeled.  173.applu models parabolic/elliptical partial differential equations and features almost exclusively floating point operations as well as a large number of memory operations making it useful as both a floating point benchmark as well as a long running memory intensive benchmark.  191.fma3d is used to model integer heavy execution.  It models finite-element crash simulation and features very few floating point operations and substantially fewer memory operations than 173.applu making it an ideal light-weight integer workload.  Finally, 183.equake is used to represent the average workload.  It models seismic wave propagation and features both integer and floating point operations as well as moderate memory usage.  As such it acts as a more representative general workload than the original default workload.  Each of the SPEC-FP benchmarks feature a 6{\char'045} register forwarding frequency as is reported by \cite{kucuk}.

\begin{table*}
\centering
\begin{tabular}{|c|c|c|c|c|} \hline
ins{\char'137}profile&default{\char'137}profile&173{\char'056}applu&191{\char'056}fma3d&183{\char'056}equake\\ \hline
loadProc&0.10&0.32&0.12&0.37\\ \hline
storeProc&0.10&0.12&0.24&0.06\\ \hline
addProc&0.10&0.01&0.15&0.08\\ \hline
subProc&0.10&0.01&0.15&0.08\\ \hline
multProc&0.10&0.01&0.15&0.08\\ \hline
divProc&0.10&0.01&0.15&0.08\\ \hline
faddProc&0.10&0.13&0.01&0.07\\ \hline
fsubProc&0.10&0.13&0.01&0.06\\ \hline
fmultProc&0.10&0.13&0.01&0.06\\ \hline
fdivProc&0.10&0.13&0.01&0.06\\ \hline
forwardingRate&0.10&0.06&0.06&0.06\\ \hline
instCount&10000&10000&10000&10000\\ \hline\end{tabular}
\caption{Benchmark Instruction Profiles showing percent rate of occurrence of each instruction type.}
\label{tab:insprofile}
\end{table*}

For the purposes of our power comparisons, each of the above instruction workloads are normalized to generate 10000 instructions.  Such a large instruction window is ideal to allow buffers to fill and behavioral characteristics to develop.  The specific values used for each workload are listed in Table ~\ref{tab:insprofile}.
\subsection{Model Implementation}
The following section details the model implementatations in our simulator.  The basic re-order buffer model we use as a base-line for comparison is a standard circular ROB model.  Section ~\ref{sec:circ} discusses its implementation.  We further model three optimizations for re-order buffers in an attempt to improve overall power usage: 1) Distributed (Cluster based) ROB, 2) Dynamically Sized ROB, 3) Reduced Output Port ROB (Retention Latch).  Each of these optimized models are built on top of the underlying circular ROB model to provide as close a behavioral and power comparison as possible.  Each of the optimizations, as well as the standard circular reorder buffer, are described in detail below.
\subsubsection{Circular ROB}
\label{sec:circ}
\subsubsection{Distributed (Cluster based) ROB}
Our distributed (cluster based) ROB is modeled similarly to the circular ROB.  Instead of having a single ROB buffer, 
numerous ROB buffers are instantiated.  In a real system, there would be one ROB buffer for each functional 
unit cluster.  However, since we do not model functional units in our simulator, we simply divide the ROB 
into multiple smaller ROBs.  Each of these ROBs implement the default circular ROB model.  Each ROB segment 
features a head pointer and a tail pointer and contains an equal portion of the total ROB space.  No further 
enhancements need to be made to the default circular ROB model as the inclusion of the central FIFO allows the 
ROB clusters to operate entirely independently of one another.  Each ROB might as well be the only ROB in the system.

The largest modeling change required by the distributed ROB is the inclusion of the central FIFO queue.  This 
FIFO queue is responsible for maintaining a global state of order between instructions in the different ROB buffers. 
The FIFO queue is modeled as a modified circular buffer and each entry contains: 1) PC of instruction, 
2) ROB_ID where the instruction placeholder resides, and 3) a valid bit corresponding to the completion state of 
the instruction.  The total number of entries in the FIFO queue is equal to the total entry size of the ROB.  
Each active entry in the FIFO queue correlates with an active entry in one of the ROB clusters.

The distributed ROB requires only minor changes to the basic circular buffer to realize its effectiveness.  Instructions are 
read from the Instruction Queue as usual and inserted into their corresponding ROB entry while updating the corresponding tail 
pointer.  Since there are now multiple ROBs into which instructions can be inserted, a quick modulus is performed on the 
destination register of the instruction to divide the instructions equally between each ROB.  This presents the possibility 
for a worst-case situation where the entire instruction stream is blocked due to improper balancing between ROB clusters.  
This situation would occur very rarely in practice and would induce no form of deadlock or livelock in the system -- 
it would merely affect performance.  Possible avoidance techniques can be researched as future work.

At the same time as inserting the instruction from the IQ into the ROB cluster, an entry must be recorded in the central FIFO 
queue to maintain global program order.  At this point, both the instruction PC and the recently determined ROB_ID are 
noted in the central FIFO queue.  The tail pointer of the central FIFO queue is increased accordingly to match.

Upon completion of an instruction, an instruction must be flagged as valid in both the ROB cluster in which it resides as well 
as its corresponding central FIFO queue entry.  The central FIFO queue then directs the appropriate ROB to write the instruction 
output to its write port when all prior instructions have been written so that program order is maintained.  This is the second 
major behavioral difference between the circular ROB and the distributed ROB, as writes and write-order is driven and determined by 
the central FIFO queue instead of the ROB cluster itself.

\subsubsection{Dynamically Sized ROB}
\subsubsection{Reduced Output Port ROB (Retention Latch)}
\subsection{Power Tabulation Implementation}
\section{Results}
hi
\subsection{subsection titl}
sub section.. with table!
\begin{table*}
\centering
\caption{Some Typical Commands}
\begin{tabular}{|c|c|l|} \hline
Command&A Number&Comments\\ \hline
\texttt{{\char'134}alignauthor} & 100& Author alignment\\ \hline
\texttt{{\char'134}numberofauthors}& 200& Author enumeration\\ \hline
\texttt{{\char'134}table}& 300 & For tables\\ \hline
\texttt{{\char'134}table*}& 400& For wider tables\\ \hline\end{tabular}
\end{table*}
% end the environment with {table*}, NOTE not {table}!

\section{Conclusions}
ending stuff
%\end{document}  % This is where a 'short' article might terminate

%ACKNOWLEDGMENTS are optional
\section{Acknowledgments}
if we really need this....

%TODO add in bib 
\begin{thebibliography}{------}
\bibitem{graphite}
	MIT Graphite Simulator
	https://github.com/mit-carbon/Graphite/wiki
\bibitem{kucuk}
Gurhan Kucuk, Dmitry Ponomarev, and Kanad Ghose. 2002. Low-complexity reorder buffer architecture. In Proceedings
of the 16th international conference on Supercomputing (ICS '02). ACM, New York, NY, USA, 57-66. DOI=10.
1145/514191.514202 http://doi.acm.org/10.1145/514191.514202
\bibitem{rupnow}
Kyle Rupnow, Arun Rodrigues, Keith Underwood, and Katherine Compton. 2006. Scientific applications vs. SPEC-FP: a comparison of program behavior. In Proceedings of the 20th annual international conference on Supercomputing (ICS '06). ACM, New York, NY, USA, 66-74. DOI=10.1145/1183401.1183413 http://doi.acm.org/10.1145/1183401.1183413 
\end{thebibliography}
\end{document}

>>>>

\maketitle
\input{start}

\input{bg}

\input{proj}

\input{end}

%\end{document}  % This is where a 'short' article might terminate

%ACKNOWLEDGMENTS are optional

%TODO add in bib
\begin{thebibliography}{------}
\bibitem{graphite}
MIT Graphite Simulator
https://github.com/mit-carbon/Graphite/wiki
\bibitem{kucuk}
Gurhan Kucuk, Dmitry Ponomarev, and Kanad Ghose. 2002. Low-complexity reorder buffer architecture. In Proceedings
of the 16th international conference on Supercomputing (ICS '02). ACM, New York, NY, USA, 57-66. DOI=10.
1145/514191.514202 http://doi.acm.org/10.1145/514191.514202
\bibitem{kucuk2}
Gurhan Kucuk, Dmitry Ponomarev, and Kanad Ghose. 2002. Energy-Efficient Design of the Reorder Buffer. In Proceedings of the International Workshop on Power and Timing Modeling, Optimization, and Simulation, September 2002.
\bibitem{kucuk3}
Gorhan Kucuk, Dmitry Ponomarev, and Kanad Ghose. 2001. Reducing Power Requirements of Instruction Scheduling Through Dynamic Allocation of Multiple Datapath Resources. In Proceedings of the International Symposium on Microarchitecture, December 2001.
\bibitem{rupnow}
Kyle Rupnow, Arun Rodrigues, Keith Underwood, and Katherine Compton. 2006. Scientific applications vs. SPEC-FP: a comparison of program behavior. In Proceedings of the 20th annual international conference on Supercomputing (ICS '06). ACM, New York, NY, USA, 66-74. DOI=10.1145/1183401.1183413 http://doi.acm.org/10.1145/1183401.1183413

\end{thebibliography}
\end{document}
